Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-07 AO_guides/manuscriptAO.tex
Last comment about forecasts and need of results from a larger scale system
second draft comments from Youyu

The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}. Forecasts from the period of 2002-2012 are available. One simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the local atmospheric pressure is also included by calculating an inverse barometer sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure set to 101,000 Pa.  The gradient of $\eta_{IB}$  is scaled by $g$ and added to the forcing terms in the ocean momentum equations \citep{madec2012nemo}. Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. Since the model domain does not include the open Pacific Ocean, pressure gradients between the open ocean and the semi-enclosed Salish Sea and their effect on the sea surface height cannot be accounted for. Previous studies in the Hudson Bay and Labrador Shelf system have argued that oscillations in the sea surface height can be explained, in part, by the response to the pressure gradient between these regions \citep{wright1987influence,young1995synoptic}.

@@

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly at the open boundary (or remote forcing). Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the remote forcing. Fourth, a simulation with only wind forcing, that is no remote forcing and no local atmospheric pressure. Comparisons between these simulations indicate the relative importance of the remote forcing, the local wind forcing, and the local atmospheric pressure forcing. Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction. Contributions due to stratification and bathymetric resolution will be left to a future study.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverse barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the effect of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation (\ref{eq:inverse}) or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer contribution is more in line with the findings of \citep{murty1995storm} but does not alter the conclusion that the remote forcing contributes most significantly. Further, the model calculates the local atmospheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. In a model domain that does not include the Pacific Ocean, this effect must be included as a boundary condition. Previous simulations by \citet{murty1995storm} suggest that the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain, consistent with the results of the present study.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge are not accounted for in the model. Finally, stratification effects on storm surge will be considered in a future study to determine if a barotropic model is reasonably accurate.

>>>>

The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}. Forecasts from the period of 2002-2012 are available. One simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

@@

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly at the open boundary (or remote forcing). Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the remote forcing. Fourth, a simulation with only wind forcing, that is, excluding remote forcing and local atmospheric pressure. Comparisons between these simulations indicate the relative importance of the remote forcing, the local wind forcing, and the local atmospheric pressure forcing. Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction. Contributions due to stratification and bathymetric resolution will be left to a future study.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. This drop is smaller than the IB sea level change calculated by equation (\ref{eq:inverse}), approximately 1 cm rise in sea level associated with 1 hPa drop in atmospheric pressure. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This magnitude is more in line with the findings of \citep{murty1995storm} that the IB contributes up to $^1/_3$ of the total surge amplitude. In our model, because the influence of remote atmospheric pressure forcing is included in the sea level at the open boundary, the above tests simply suggests that the local atmopsheric pressure generates a sea level changes that is smaller than the IB response. This is related to the small spatial scale of the model domain.

Other findings have reported the inverse barometer effect due to atmospheric pressure contributes \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverse barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the effect of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation (\ref{eq:inverse}) or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. Further, the model calculates the local atmospheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

These comparisons point to the importance of including, as a forcing condition at the open boundary, the surge entering the domain from the Pacific Ocean. Storms travelling over the Pacific Ocean towards the west coast of the northern United States and Canada can induce elevated sea levels along the coastline which then enter the Salish Sea system through the Strait of Juan de Fuca. Previous simulations by \citet{murty1995storm} suggest that the surge entering the system through the Strait of Juan de Fuca contribute most significantly to storm surges in this domain, consistent with the results of the present study.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge are not accounted for in the model. Since the remote forcing is an important contributing factor, any forecasting with this regional model must make use of results from a larger scale prediction system. Finally, stratification effects on storm surge will be considered in a future study to determine if a barotropic model is reasonably accurate.
